---
title: "An intro to R with an Epi working example"
author: '[Crippa Alessio](https://alecri.github.io/)'
date: "December 10th, 2019"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Aim

The aim of this document is to give a short introduction to the R software. During
this tutorial I will briefly cover how to:  

- read, explore and manipulate data;  
- obtain summary statistics and produce informative graphics;  
- fit statistical models.  

The topics will be covered using the collection of packages in [tidyverse](https://www.tidyverse.org/). 
The real data from a cohort of marathon runners will serve as working example.

**Reference**  
[Hyponatremia among Runners in the Boston Marathon, *New England Journal of Medicine*, 2005, Volume 352:1550-1556](https://www.nejm.org/doi/full/10.1056/NEJMoa043901).  

**Descriptive abstract**  
Hyponatremia has emerged as an important cause of race-related death and life-threatening 
illness among marathon runners. We studied a cohort of marathon runners to estimate the 
incidence of hyponatremia and to identify the principal risk factors.

## Load R packages

The following packages are used in the tutorial. If you want to reproduce the code
chunks below, be sure to install them (e.g. `install.packages("tidyverse")`).

```{r}
library(tidyverse)
library(Epi)
library(knitr)
library(gridExtra)
```


## Read, explore and manipulate data

### Read data

The `marathon` data are available in a csv format at http://alecri.github.io/downloads/data/marathon.csv .

```{r}
marathon <- read_csv("http://alecri.github.io/downloads/data/marathon.csv")
marathon
```

Many other options exist, such as:  

- basic functions (`read.table`, `read.csv`, `read.delim`)  
- from Excel `readxl::read_excel`  
- from SPSS `haven::read_sav`  
- from SAS `haven::read_sas`
- from Stata `haven::read_dta`

A comprehensive tutorial can be found [here](https://www.datacamp.com/community/tutorials/r-data-import-tutorial).

### Explore the data

```{r}
# what is the dimension of the data  
dim(marathon)
# i.e. how many obs/rows
nrow(marathon)
# how many cols/variables?
ncol(marathon)
# what's the name of the variable
colnames(marathon)
# what is the structure of the data
glimpse(marathon)
# show the first rows
head(marathon)
# quick and dirty summary
summary(marathon)
```


### Manipulate data

Data manipulations consist of a multiplicity and combinations of steps.
The `dplyr` package in `tidyverse` aids this task with a collections of *verbs*.  
I will show how to:  

1. create/modify variables with the `mutate` verb.  

```{r}
# run time in hour
marathon <- mutate(marathon, runtime_h = runtime/60)
# alternative code
# marathon$runtime_h <- marathon$runtime/60
```

```{r}
# modify a character variable into a factor variable
str(marathon$nas135)
marathon <- mutate(marathon, nas135 = factor(nas135, levels = c("na > 135", "na <= 135")))
str(marathon$nas135)
```

```{r}
# categorize a continuous variable
summary(marathon$age)
marathon <- mutate(marathon, age_cat = cut(age, breaks = c(19, 31, 38, 45, 75)))
table(marathon$age_cat)
```

2. select only variables of interest with the `select` verb.

```{r}
# some variables + variables containing wt (shortage for weight)
marathon_sub <- select(marathon, id, na, female, bmi, runtime_h, contains("wt"))
head(marathon_sub)
```

3. select only some observations with the `filter` verb.

```{r}
female_27 <- filter(marathon_sub, female == "female", bmi > 27)
female_27
```

4. sort data according to increasing/decreasing values of some variables with the
`arrange` verb.

```{r}
arrange(female_27, desc(na), bmi)
```

Chaining: wrap different functions inside each other

```{r}
arrange(
   filter(
      select(
         mutate(marathon,
                runtime_h = runtime/60),
         id, na, female, bmi, runtime_h, contains("wt")),
      female == "female", bmi > 27), desc(na), bmi)
```

#### The %>% (pipe) operator  
This operator allows you to pipe the output from one function to the input of another function.

```{r}
marathon %>% 
   mutate(runtime_h = runtime/60) %>%
   select(id, na, female, bmi, runtime_h, contains("wt")) %>%
   filter(female == "female", bmi > 27) %>%
   arrange(desc(na), bmi)
```

## Summary statistics and graphs

### ggplot2: grammar of graphics

```{r}
ggplot(marathon, aes(wtdiff, na, col = bmi)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ female) +
  labs(x = "Weight change, kg", y = "Sodium concentration, mmol per liter") +
  theme_bw()
# change theme for next graphs
theme_set(theme_bw())
```


### Univariate descriptives

```{r}
# for a continuous variable
summary(marathon$na)
c(mean = mean(marathon$na), sd = sd(marathon$na))
quantile(marathon$na)
summarise(marathon, mean = mean(na), sd = sd(na), median = median(na), iqr = IQR(na))
# more concise
summarise_each(marathon, funs(mean, sd, median, IQR), na)
# histogram (counts)
ggplot(marathon, aes(na)) +
  geom_histogram() +
  labs(x = "Sodium concentration, mmol per liter")
# histogram (density)
ggplot(marathon, aes(na)) +
  geom_histogram(stat = "density", n = 2^5) +
  geom_line(stat = "density") +
    labs(x = "Sodium concentration, mmol per liter")
# boxplot
ggplot(marathon, aes(x = "", y = na)) +
  geom_boxplot() + 
  labs(x = "", y = "Sodium concentration, mmol per liter")
```

```{r}
# for a categorical variable
table(marathon$nas135)
prop.table(table(marathon$nas135))
# barplot (counts)
ggplot(marathon, aes(nas135)) +
  geom_bar()
# barplot (proportion)
ggplot(marathon, aes(nas135)) +
  geom_bar(aes(y = ..count../sum(..count..))) +
  labs(x = "", y = "Prevalence of hyponatremia")
```


### Bivariate descriptives (association)

```{r}
# Continuous + categorical
marathon %>% 
  group_by(female) %>% 
  summarise_each(funs(mean, sd, median, IQR), na)
ggplot(marathon, aes(female, na)) +
  geom_boxplot() +
  labs(x = "", y = "Sodium concentration, mmol per liter")
ggplot(marathon, aes(female, na, fill = female)) +
  geom_violin() +
  guides(fill = "none") +
    labs(x = "", y = "Sodium concentration, mmol per liter")
ggplot(marathon, aes(na, col = female)) +
  geom_line(stat = "density") +
  labs(x = "Sodium concentration, mmol per liter", col = "Sex")
# statistical tests
t.test(na ~ female, data = marathon, var.equal = FALSE)
wilcox.test(na ~ female, data = marathon)
```

```{r}
# 2-by-2 table
tab1 <- with(marathon, table(nas135, female))
tab1
prop.table(tab1, margin = 2)
ggplot(marathon, aes(nas135, fill = female)) +
  geom_bar(position = "dodge")
ggplot(marathon, aes(nas135, group = female, fill = female)) +
  geom_bar(aes(y = ..prop..), position = "dodge")
# statistical tests
chisq.test(tab1)
```

```{r}
# 2 continuous
cor_nawtd <- with(marathon, cor(na, wtdiff,  method = "pearson", use = "complete.obs"))
cor_nawtd
p1 <- ggplot(marathon, aes(wtdiff, na)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Weight change, kg", y = "Sodium concentration, mmol per liter")
library(ggExtra)
ggMarginal(p1, type = "histogram")
```

#### Table 1

```{r}
library(table1)
table1(~ age + female + wtdiff + bmi + runtime | nas135, data = marathon)
```


## Statistical models

### Linear regression

$$E[Y|X] = \beta_0 + \beta_1X_1 + \dots + \beta_kX_k$$

```{r}
# univariate linear regression
fit1 <- lm(na ~ female, data = marathon)
summary(fit1)
ci.lin(fit1) %>% kable()
```

```{r}
# multivariable linear regression
fit2 <- lm(na ~ wtdiff + female, data = marathon)
ci.lin(fit2) %>% kable()
marathon <- mutate(marathon,
                   pred_fit2 = predict(fit2, newdata = marathon))
```

```{r}
# multivariable linear regression with interaction
fit3 <- lm(na ~ wtdiff*female, data = marathon)
ci.lin(fit3, ctr.mat = rbind(c(0, 1, 0, 0),
                             c(0, 1, 0, 1)))
marathon <- mutate(marathon,
                   pred_fit3 = predict(fit3, newdata = marathon))
```

```{r}
# graphical comparison
grid.arrange(
  ggplot(marathon, aes(wtdiff, pred_fit2, col = female)) +
    geom_line() +
    labs(x = "Weight change, kg", y = "Predicted mean na"),
    ggplot(marathon, aes(wtdiff, pred_fit3, col = female)) +
    geom_line() +
    labs(x = "Weight change, kg", y = "Predicted mean na"),
  ncol = 2
)
```


### Logistic regression

$$\log(\textrm{odds}(Y|X)) = \beta_0 + \beta_1X_1 + \dots + \beta_kX_k$$
```{r}
# univariate logistic regression
fit4 <- glm(nas135 ~ wtdiff, data = marathon, family = binomial)
ci.exp(fit4)
marathon <- marathon %>% 
  mutate(pred_p = predict(fit4, newdata = marathon, type = "response"),
         pred_odds = predict(fit4, newdata = marathon, type = "link")) 
grid.arrange(
  ggplot(marathon, aes(wtdiff, pred_p)) +
    geom_line() +
    labs(x = "Weight change, kg", y = "Predicted probability"),
  ggplot(marathon, aes(wtdiff, pred_odds)) +
    geom_line() +
    labs(x = "Weight change, kg", y = "Predicted odds"),
  ncol = 2
)
```
```{r}
# multivariable logistic regression
fit5 <- glm(nas135 ~ wtdiff + female, data = marathon, family = binomial)
ci.exp(fit5)
mutate(marathon, pred = predict(fit5, newdata = marathon, type = "response")) %>% 
  ggplot(aes(wtdiff, pred, col = female)) +
  geom_line() +
  labs(x = "Weight change, kg", y = "Predicted probability")
```
